---
title: "Iris plotting example"
output: 
  html_document: 
    number_sections: yes
    toc: yes
---

# Libraries used

```{r, eval=FALSE}
install.packages(c("broom", "xtable", "gee", "RColorBrewer", "gplots"))
```

# Exploring the Iris dataset

## Basic introduction to the dataset:
```{r}
data(iris)
dim(iris)
class(iris)
summary(iris)
head(iris)
sapply(iris, class)
```

## Plotting
Default plot for a numeric variable:
```{r}
plot(iris$Sepal.Length)
```

For two numeric variables:
```{r}
plot(iris$Sepal.Length, iris$Sepal.Width)
```

Equivalently using the formula interface:
```{r, eval=FALSE}
plot(iris$Sepal.Width ~ iris$Sepal.Length)
```

Or:
```{r, eval=FALSE}
pdf("iris1.pdf")
plot(Sepal.Width ~ Sepal.Length, data=iris)
dev.off()
```

## Details in plotting

Make it look a better. Look at ?plot and ?plot.default.  Play with type, ylim, log, main, sub, xlab, ylab, pch, cex.  Explore different output devices, and RStudio "Export."

Do again, using different shapes (pch) or colors (col) for each species.  Add a legend.

```{r}
plot(1:20, pch=1:20)
plot(1:20, pch=c(".", "a", "dog"))
```

## Color palettes.
Several package provide color palettes:
```{r}
library(RColorBrewer)
display.brewer.all(n=3)
brewer.pal(3, "Set2")
```

You can also create your own:
```{r}
plot(1:10, col=colorRampPalette(c("red", "lightgrey", "black"))(10), pch=15, cex=3)
```

**Exercise** try creating and using some different color palettes

```{r}
plot(Sepal.Width ~ Sepal.Length, data=iris, 
     col=brewer.pal(3, "Set2")[iris$Species])
legend(x=6.5, y=4.5, legend=levels(iris$Species), 
       col=brewer.pal(3, "Set2"), pch=1)
```

**Note:** how did I know to do x$mids, x$counts?
```{r}
x <- hist(iris$Sepal.Length)
segments(x0=x$mids-0.25, x1=x$mids+0.25, y0=x$counts, y1=x$counts, lw=4, col="red")
```

How about a boxplot with histogram together? (see http://rgraphgallery.blogspot.com/search/label/boxplot)
```{r}
layout(mat = matrix(c(1,2), 2, 1, byrow=TRUE),  height = c(1,3))
par(mar=c(2, 2, 0.5, 1))
boxplot(iris$Sepal.Length, horizontal=TRUE,  outline=TRUE,ylim=c(4,8), frame=FALSE, col = "green1")
hist(iris$Sepal.Length,xlim=c(4,8), col = "pink", freq = FALSE)
lines(density(iris$Sepal.Length))
```

Boxplot of Sepal width vs. species
```{r}
boxplot(Sepal.Width ~ Species, data=iris, boxwex=0.25, col="grey")
```

```Use dev.off() to clear layout.```

```{r, fig.scap="Heatmap of iris dimensions.", message=FALSE, warning=FALSE}
library(gplots)
heatmap.2(t(iris[, 1:4]), trace="none", scale="row", 
          key=TRUE, mar=c(2, 8), cexRow=1, 
          ColSideColors=c("grey", "black", "yellow")[iris$Species])
```

See https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html for more advanced and interactive heatmaps.

Exploratory pairs plots:
```{r}
pairs(iris)
```

## Final notes on plotting

* The ggplot2 package provides nicer-looking plots by default
* There is a lot to learn: https://mramos.shinyapps.io/obView

# Plotting and exploratory data analysis exercises

Datasets are available at http://goo.gl/x87GY6

## Exercise #1: Iris dataset

Plot petal width vs. petal length for the iris dataset, identifying species using _both_ color and datapoint type.
```{r}
data(iris)
plot(Petal.Width ~ Petal.Length, col=c("black", "red", "blue")[Species], pch=(15:17)[Species], data=iris)
legend("topleft", legend=levels(iris$Species), col=c("black", "red", "blue"), pch=15:17)
```


## Exercise #2: GSE12945

Do exploratory analysis of the GSE12945 dataset, taken from the following paper:

> Staub E, Groene J, Heinze M, Mennerich D et al. An expression module of WIPF1-coexpressed genes identifies patients with favorable prognosis in three tumor types. J Mol Med (Berl) 2009 Jun;87(6):633-44. PMID: 19399471

Dataset is available from http://goo.gl/x87GY6 or by using the following command: 
```{r}
if(!file.exists("GSE12945.csv"))
  download.file("https://www.dropbox.com/sh/pukanjaahmonmcp/AADWX-vKk70CuGgYWBqqxWjfa/datasets/GSE12945.csv", destfile="GSE12945.csv", method="wget")
gse <- read.csv("GSE12945.csv")
```

# Regression

## Simple Linear Regression Example:
```{r}
data(iris)
fit <- lm(Petal.Width ~ Petal.Length, data=iris)
class(fit)
methods(class=class(fit))
summary(fit) # show results
```

## Other useful functions 
```{r, results='hide'}
coefficients(fit) # model coefficients
predict(fit) # fitted predictions
predict(fit, newdata=data.frame(Petal.Length=seq(1, 2, by=0.1)))
confint(fit, level=0.95) # CIs for model parameters 
fitted(fit) # predicted values
residuals(fit) # residuals
anova(fit) # anova table 
influence(fit) # regression diagnostics
```

The "broom" package can be useful for extracting key results:
```{r}
library(broom)
tidy(fit)
```

## Diagnostic plots
```{r}
par(mfrow=c(2,2))
plot(fit)
```

## Multiple regression
```{r}
fit2 <- lm(Petal.Width ~ Petal.Length + Sepal.Length + Sepal.Width, data=iris)
summary(fit2) # show results
anova(fit, fit2)
```

## Interaction terms
```{r}
fit2int <- lm(Petal.Width ~ Petal.Length + Sepal.Length + Sepal.Width + Petal.Length:Sepal.Length, data=iris)
anova(fit2, fit2int)
```

## Model matrices
```{r}
tail(model.matrix(~ Species, data=iris))
tail(iris)
```

## Analysis of covariance
```{r}
fit3 <- lm(Petal.Width ~ Petal.Length + Sepal.Length + Sepal.Width + Species, data=iris)
summary(fit3)
anova(fit2, fit3)
```

Make nice tables - see http://cran.r-project.org/web/packages/xtable/vignettes/xtableGallery.pdf.

```{r, results='asis'}
library(xtable)
print(xtable(fit3), type="html")
print(xtable(anova(fit3)), type="html")
```

## Logistic regression
```{r}
iris2 <- iris
iris2$virginica <- iris$Species == "virginica"
fit4 <- glm(virginica ~ Petal.Width + Petal.Length + Sepal.Length + Sepal.Width, data=iris2, family=binomial)
plot(fit4)
```

## Poisson regression

(obviously a bad choice here):
```{r}
fit5 <- glm(round(Petal.Width) ~ Petal.Length + Sepal.Length + Sepal.Width, data=iris2, family=poisson)
plot(fit5)
summary(fit5)
```

## Generalized Estimating Equations
```{r, eval=FALSE}
library(gee)
gee(formula, id, data, ...)
```

# Regression exercises
## Exercise #1
Plotting a linear regression with confidence and prediction intervals.

```{r}
plot(Petal.Width ~ Petal.Length, col=c("black", "red", "blue")[Species], pch=(15:17)[Species], xlab="Petal Length (cm)", ylab="Petal Width (cm)", data=iris)
newx <- data.frame(Petal.Length=seq(min(iris$Petal.Length), max(iris$Petal.Length), length.out=100))
conf.interval <- predict(fit, newdata=newx, interval="confidence")
pred.interval <- predict(fit, newdata=newx, interval="prediction")
lines(conf.interval[, "fit"] ~ newx[, 1], lty=1, lw=3)
lines(conf.interval[, "lwr"] ~ newx[, 1], lty=2)
lines(conf.interval[, "upr"] ~ newx[, 1], lty=2)
lines(pred.interval[, "lwr"] ~ newx[, 1], lty=3)
lines(pred.interval[, "upr"] ~ newx[, 1], lty=3)
legend("topleft", legend=c(levels(iris$Species), "CI", "PI"), col=c("black", "red", "blue", "black", "black"), pch=c(15:17, -1, -1), lty=c(-1, -1, -1, 2, 3))
```

## Exercise #2
Regression on the GSE12945 dataset.  Is there an association between WIPF1 and any of the other variables?  Note: this exercise is meant to do interactively with the class, otherwise the code will look too advanced for beginners.

```{r}
download.file("https://www.dropbox.com/sh/pukanjaahmonmcp/AADWX-vKk70CuGgYWBqqxWjfa/datasets/GSE12945.csv", destfile="GSE12945.csv", method="wget")
gse <- read.csv("GSE12945.csv", row.names=1)  #note that row.names is needed in this case to make the first column row names.
sapply(gse, class) #make sure classes are appropriate
```

G, T, N, M should be ordered factors:
```{r}
gse$G <- factor(gse$G, ordered=TRUE)
gse$T <- factor(gse$T, ordered=TRUE)
gse$N <- factor(gse$N, ordered=TRUE)
gse$M <- factor(gse$M, ordered=TRUE)
```

This is one way to select all columns except for WIPF1:
```{r}
(vars.to.test <- colnames(gse)[!colnames(gse) %in% "WIPF1"])
```

```{r}
all.lm <- lapply(vars.to.test, function(x){
  lm(as.formula(paste("WIPF1 ~", x)), data=gse)
})
names(all.lm) <- vars.to.test
all.p <- sapply(all.lm, function(mylm){
  fstat <- summary(mylm)$fstatistic
  1-pf(fstat[1], fstat[2], fstat[3])
})
t(all.p)
hist(all.p)
```

It doesn't look like it!
