---
title: "Making the most of R"
author: "Marcel Ramos"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  beamer_presentation:
    theme: "CambridgeUS"
    colortheme: "dolphin"
    fonttheme: "structureitalicserif"
---

Using R for Big Data
==============================================
* Big Data 
* Tidy Data
* Tips for learning R
* Reading data into memory
* Cleaning and Manipulating data with `tidyr` and `dplyr`
* Pipes for fluid and readable programming

Big Data\footnote{\tiny{Slide adapted from Hadley Wickham}}
==============================================
Size   | Description
------ | -------
Big    | Can't fit in memory on one computer: >5 TB
Medium | Fits in memory on a server: 10 GB - 5 TB
Small  | Fits in memory on a laptop: <10 GB

\begin{block}{Note:}
R is great at small!
\end{block}

Big Data (2)
==============================================
* Reducible problems (subsetting, sampling, summarizing)
* Big data is often messy data and not much else
* Price to pay for big data

Principles of Tidy Data
==============================================
* Often said: 80% of data analysis is cleaning/munging
* Provide a standard way of organizing data\footnote{\tiny{http://vita.had.co.nz/papers/tidy-data.pdf}}
1. Each variable forms a column
2. Each observation forms a row
3. Each type of observational unit forms a table

Dataset     | Variable    | Variable
----------- | ----------- | -----------
Observation | Value       | Value
Observation | Value       | Value


Principles of Tidy Data (2)
==============================================
* Why is tidy data important? 
* Easier for the analyst and the computer to extract knowledge from a set of values
* Saves a *lot* of time

Tips for learning R (general)
==============================================
* Learning R may become frustrating at times
* Learning a language
* Practice is key

Useful tips for learning R (stand-alone)
==============================================
Pseudo code                                   |   Example code
--------------------------------------------  |   -------------------
install.packages(packagename)                 | library(dplyr) 
?functionname                                 | ?select
?package::functionname                        | ?dplyr::select
? 'Reserved keyword or symbol' \color{blue}{(or backticks)} | ? '%>%'
??searchforpossiblyexistingfunctionandortopic | ??simulate
help(package = "loadedpackage")               | help("dplyr")
browseVignettes("packagename")                | browseVignettes("dplyr")

Learning R via online courses
==============================================
* [Coursera](https://www.coursera.org)
* [edX](https://www.edX.org)
* [RStudio](https://www.rstudio.com/resources/training/online-learning/#R)
* [Quick-R](http://www.statmethods.net/) -
Mostly for basic and base functions
* [RStudio Cheatcheets](https://www.rstudio.com/resources/cheatsheets/)

Reading and Loading Datasets into Memory
==============================================
* Requires installation of `devtools` package and Rtools (varies by OS)
```{r, eval = FALSE, echo = TRUE}
devtools::install_github("username/repository")
devtools::install_github("hadley/readr")
devtools::install_github("hadley/haven")
```

Read Time
==============================================
```{r, eval = FALSE, echo = TRUE}
file.info("data/BRFSS2013_Data.csv")$size/(1024^2)
system.time(read.csv("data/BRFSS2013_Data.csv"))

library(readr)
system.time(read_csv("data/BRFSS2013_Data.csv"))
```

58.8 MB File

Function         | Elapsed Time
---------------- | ------------
utils::read.csv  | 5.115
readr::read_csv  | 1.836

Read Time (2)
==============================================

* You may also consider the `fread` function
* data.table syntax is different

```{r, eval = FALSE, echo = TRUE}
library(data.table)
?fread
```

Data Munging using `tidyr`
==============================================
* `tidyr` faciliates reshaping of data
1. spread vs. gather \color{purple}{*most likely to use}\color{black}
2. extract/separate vs. unite
3. nest vs. unnest

Data Manipulation using `dplyr`
==============================================
* `dplyr` convention aims to ease cognitive burden
* Function names are easy to remember
1. select (Y)
2. mutate/transmute (add Ys / new Y)
3. filter (get Xs based on condition)
4. slice (get Xs specified)
5. summarise (reduce to single observation) 
6. arrange (re-order observations)

Examples of use
==============================================
* Create an example of messy data: 

```{r, echo = TRUE, message = FALSE}
library(dplyr); library(tidyr)
data("mtcars")
mtcars <- select(mtcars, c(mpg:hp, wt, vs:carb))
mtcars <- unite(mtcars, cylgear, cyl, gear)
separate(mtcars, cylgear, c("cyl0", "gear0"))[1:3,]
```

mtcars <- select(mtcars, c(1:4, 5, 7:11))

Mutate & Transumte
==============================================
```{r, echo = TRUE, message = FALSE}
head(mutate(mtcars, displ_l = disp/61.0237))
head(transmute(mtcars, disp_l = disp/61.0237),2)
```

Example with base functions
==============================================
```{r, eval = TRUE, echo = TRUE}
data("mtcars")
mtcars <- mtcars[,c("mpg", "cyl", "disp", "hp",
                    "wt", "vs", "am", "gear", "carb")]
mtcars$cylgear <- with(mtcars, paste(cyl, gear, sep = "."))
mtcars[, c("cyl1", "gear1")] <- NA
mtcars[, c("cyl1", "gear1")] <-
  t(sapply(strsplit(mtcars$cylgear, ".", fixed = TRUE), FUN = "[", c(1,2)))
head(mtcars, 3)
```

Considerations
==============================================

\huge{Be careful of loss of information!}

Functional programming example
==============================================
```{r, eval = FALSE, echo = TRUE}
hourly_delay <- filter(
  summarise(
    group_by(
      filter(
        flights,
        !is.na(dep_delay)
      ),
      date, hour
    ), 
    delay = mean(dep_delay),
    n = n()
  ), 
  n > 10
)
```

Pipes for fluid and readable programming
==============================================
* Piping operator: %>%
* Consider the previous example with pipes:

```{r, echo = TRUE, eval = FALSE}
hourly_delay <- flights %>% 
  filter(!is.na(dep_delay)) %>%
  group_by(date, hour) %>%
  summarise(delay = mean(dep_delay), n = n()) %>%
  filter(n > 10)
```

More piping
==============================================
```{r, echo = TRUE, message = FALSE}
library(nycflights13)
flights %>% group_by(carrier) %>% 
  summarise(avg_depdelay = mean(dep_delay, na.rm = TRUE), count = n()) %>% 
  left_join(airlines) %>% arrange(avg_depdelay) %>% head
```

Using `separate`
==============================================
```{r , echo = TRUE, message = FALSE}
data(iris)
longdata <- gather(iris, key = measure, n, Sepal.Length:Petal.Width) %>% 
  separate(measure, c("type", "dimension"))
longdata %>% group_by(Species, type, dimension) %>% summarise(avg_dim = mean(n, na.rm = TRUE))
```

Piping with `tidyr`
==============================================
```{r, echo = TRUE}
library(readr)
(pew <- read_csv("../data/pew.csv"))
```

Using `gather`
==============================================

```{r, echo = TRUE}
pew %>% gather(income, n, -religion) %>% head
```

income, religion : variables to gather 
n : variable in cells 
-religion means all except religion

Using `group_by`
==============================================
```{r, echo = TRUE}
pew %>% gather(income, n, -religion) %>% group_by(income) %>% summarise(totals = sum(n))
```

Using `group_by` (2)
==============================================
```{r, echo = TRUE}
pew %>% gather(income, n, -religion) %>% group_by(religion) %>% summarise(totals = sum(n))
```

Summary
==============================================
* Big data not always the best option
* Tidy data makes everything easier and saves time
* Learning R can be a bit frustrating but certainly not impossible
* R is great for small types of datasets that fit into memory but can also be used in HPC
* Writing R code should not be a cognitive burden on the user 
* R programming should be readable and fun to use!
